A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-03-20 paper/paper.tex
Version of the response sent

As a result, our work can be seen as contributing the following: it characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions; %Second, it makes fully formal the relation between semi-simplicial and semi-cubical sets and parametricity.
it provides with an alternative case study in the realm of indexed constructions of semi-simplicial and semi-cubical sets~\cite{voevodsky12,herbelin15,part15}, especially in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\cite{part15,shulman14,altenkirch16,kraus2017space}, as well as of better understanding the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}); it makes a step in the direction of the program initiated in~\cite{altenkirch15} to develop parametricity-based models of parametric type theory~\cite{bernardy15,nuyts17,cavallo19} and cubical type theory~\cite{bezem13,cohen16,angiuli21} which are closer from the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do (compare e.g. to the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally closed cartesian categories~\cite{curien14}).

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we build up intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

We give in this section a uniform definition of semi-simplicial and semi-cubical sets which we believe is folklore. The uniformity allows the definition of a notion of $\nu$-sets that subsumes semi-simplicial and semi-cubical sets.

@@

The definition is dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and~\ref{tab:coh}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a stream representing the (coinductive) limit of $n$-truncated $\nu$-sets. The $n$-truncated $\nu$-sets are themselves described in \ref{tab:core}. In such stream, the $n$th component is a type dependent over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.

>>>>

As a result, our work can be seen as contributing the following: it characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions; %Second, it makes fully formal the relation between semi-simplicial and semi-cubical sets and parametricity.
it provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\cite{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\cite{part15,shulman14,altenkirch16,kraus2017space}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}); it makes a step in the direction of the program initiated in~\cite{altenkirch15} to develop parametricity-based models of parametric type theory~\cite{bernardy15,nuyts17,cavallo19} and cubical type theory~\cite{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do (compare e.g. to the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~\cite{curien14}).

The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. We then proceed to explain the connection between the indexed representation and parametricity in \ref{sec:indexed-param}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

@@

In this section, we give a uniform definition of semi-simplicial and semi-cubical sets which we believe is folklore. The uniformity allows the definition of a notion of $\nu$-sets that subsumes semi-simplicial and semi-cubical sets.

@@

The definition is dispatched over tables \ref{tab:coind}, \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and~\ref{tab:coh}. Table \ref{tab:coind} describes a $\nu$-set in indexed form, as a stream, coinductively representing the limit of $n$-truncated $\nu$-sets. The $n$-truncated $\nu$-sets are themselves described in \ref{tab:core}. In such stream, the $n$th component is a type dependent over a $\fullframe$. The type $\fullframe$ is recursively defined in \ref{tab:frames}, using the auxiliary definitions of $\layer$ and $\painting$. The type $\layer$ and $\painting$ are dependent over $\restrf$, and these restrictions are defined on \ref{tab:faces}. These restrictions are defined using auxiliary definitions of $\restrl$ and $\restrc$.
